---
title: "S2 - Read and data plotting from the Manuscript Marreiros et al. 2022"
author: "Joao Marreiros, Mae Goder-Goldberger, Eduardo Paixao"
date: "7/21/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(comment = NA, message = FALSE, indent = "", error = TRUE)

```

__Brief description of the script__

This R markdown document reads, summarizes and plots data concerning the analysis of the Manot core engraved incisions, included in the manuscript Marreiros et al. 2022.

The document contains:

1. Tables
2. Figures (data analysis)

This R project and respective scripts follow the procedures described by Marwick et al. 2017. 

To compile this markdown document do not delete or move files from their original folders.
Please note that most of the tables and figures in this file do not match the numbering in the original manuscript.

For any questions, comments and inputs, please contact:

João Marreiros, joao.marreiros@rgzm.de


# Load data into R project

*Imported files are in: '`r paste0("../analysis/raw_data")`'*

*Figures are saved in: '`r paste0("../analysis/plots")`'*

*Tables are saved in: '`r paste0("../analysis/derived_data")`'*

# Load libraries

```{r}

library(tidyverse)
library(knitr)
library(kableExtra)
library(GGally)
library(doBy)
library(ggfortify)
library(tools)
library(rstatix)
library(ggpubr)

```


# Import dataset

```{r}

db <- read_csv("../raw_data/incisionsangle.csv", na = c("", "NA"), quoted_na = TRUE)
data_file <- list.files("../raw_data/", pattern = "\\.csv$", full.names = TRUE)

```

Show dataset:

```{r}

str(db)

```

# Import and summarize data

```{r fig.height = 6.5, fig.width = 8.5, res = 300}

# compute descriptive statistics

nminmaxmeanmedsd <- function(x){
	y <- x[!is.na(x)]
	n_test <- length(y)
	min_test <- min(y)
	max_test <- max(y)
	mean_test <- mean(y)
 	med_test <- median(y)
 	sd_test <- sd(y)
 	out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test)
 	names(out) <- c("n", "min", "max", "mean", "median", "sd")
 	return(out)
}

num.var <- 5:length(db)

anglestats <- summaryBy(.~aoi + site + incision, data=db[c("site", "incision", names(db)[num.var])], FUN=nminmaxmeanmedsd)

anglestats

write_csv(anglestats, "../summary_stats/anglestats.csv")

```

# Plot data

```{r fig.height = 6.5, fig.width = 8.5, res = 300}

# converting to categorical
db$aoi = as.factor(db$aoi) 
# db$incision = as.factor(db$incision)

sitesangle <- ggplot(data = db, aes(x = site, y = angle)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs( x = "", y = "Angle in º") +
           geom_jitter(aes(colour = site), alpha=0.9, position=position_jitter(w=0.1,h=0.1))

print(sitesangle)

ggsave("../plots/sitesangle.png")


```

# Run ANOVA for comparison

```{r}



# brief visualization of the data from the 3 artifacts

ggplot(data = db, aes(x= site, y= angle)) +
  geom_boxplot()

# check for outliers

db %>%
  group_by(site) %>%
  identify_outliers(angle)

# there are no extreme outliers

# check normality

model <- lm(angle ~ site, data = db)

# QQ plot of residuals

ggqqplot(residuals(model))
shapiro_test(residuals(model))

db %>%
  group_by(site) %>%
  shapiro_test(angle)

ggqqplot(db, "angle", facet.by = "site")

ang.aov <- db %>% anova_test(angle ~ site)
ang.aov

# post-hoc test

ang.pwc <- db %>% tukey_hsd(angle ~ site)
ang.pwc

# plot the results

ang.pwc.plot <- ang.pwc %>% add_xy_position(x = "site")


ggboxplot(db, x = "site", y = "angle") +
  stat_pvalue_manual(ang.pwc.plot, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(ang.aov, detailed = TRUE),
    caption = get_pwc_label(ang.pwc.plot)
    )

```

# Kruskal-Wallis test
```{r}

ang.kruskal <- db %>% kruskal_test(angle ~ site)
ang.kruskal

db %>% kruskal_effsize(angle ~ site)

# comparison between pairs

ang.will <- db %>% 
  wilcox_test(angle ~ site, p.adjust.method = "bonferroni")
ang.will

ang.will <- ang.will %>% add_xy_position(x = "site")

ggboxplot(db, x = "site", y = "angle") +
  stat_pvalue_manual(ang.will, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(ang.kruskal, detailed = TRUE),
    caption = get_pwc_label(ang.will)
    )

ggsave("../plots/anova.png")

```


---

End of Script










