---
title: "Reads and plots data from the Manuscript Goder-Goldberger et al AASc, In press"
author: "Mae Goder-Goldberger, Joao Marreiros, Eduardo Paixao, Erella Hovers"
date: "`r Sys.time()`"
output:
  html_document: default
  pdf_document: default
---

__Brief description of the script__

# Content
This script reads and plots the data from the 3D and 2D morphometric analysis of the artefacts' engraved incisions. For details on the methods and data acquisition, please visit the Materials and Methods section of the paper. 

The knit directory for this script is the project directory.This R project and respective scripts follow the procedures described by Marwick et al. 2017. 

For any questions, comments and inputs, please contact:

João Marreiros, joao.marreiros@leiza.de

# Load packages
```{r}

library(R.utils)
library(ggplot2)
library(tools)
library(tidyverse)
library(doBy)
library(ggrepel)
library(flextable)
library(readr)
library(ggpubr)
library(crosstable)
library(car)

```


# Import and preview raw data
```{r}

db <- read.csv("../raw_data/data.csv", stringsAsFactors=FALSE, sep=";", dec=".")

str(db)
head(db)

```

# Import and summarize data

```{r}

# summarise data frome ach artefact, organised by AOI

# sample data per artefact

manotdb <- filter(db, artefact == "Manot")
amud1db <- filter(db, artefact == "Amud1")
amud2db <- filter(db, artefact == "Amud2")
qzdb <- filter(db, artefact =="Qafzeh")
qnrtdb <- filter(db, artefact =="Quneitra")

# Manot
# Horizontal distance
manotdistance <- manotdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
manotdepth <- manotdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
manotarea <- manotdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
manotangle <- manotdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
manotdistance
manotdepth
manotarea
manotangle

# save the results 
write_csv(manotdepth, "../summary_stats/manotdepth.csv")
write_csv(manotarea, "../summary_stats/manotarea.csv")
write_csv(manotangle, "../summary_stats/manotangle.csv")
write_csv(manotdistance, "../summary_stats/manotdistance.csv")

# Amud1
# Horizontal distance
amud1distance <- amud1db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
amud1depth <- amud1db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
amud1area <- amud1db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
amud1angle <- amud1db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
amud1distance
amud1depth
amud1area
amud1angle

# save the results 
write_csv(amud1depth, "../summary_stats/amud1depth.csv")
write_csv(amud1area, "../summary_stats/amud1area.csv")
write_csv(amud1angle, "../summary_stats/amud1angle.csv")
write_csv(amud1distance, "../summary_stats/amud1distance.csv")

# Amud2
# Horizontal distance
amud2distance <- amud2db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
amud2depth <- amud2db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
amud2area <- amud2db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
amud2angle <- amud2db %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
amud2distance
amud2depth
amud2area
amud2angle

# save the results 
write_csv(amud2depth, "../summary_stats/amud2depth.csv")
write_csv(amud2area, "../summary_stats/amud2area.csv")
write_csv(amud2angle, "../summary_stats/amud2angle.csv")
write_csv(amud2distance, "../summary_stats/amud2distance.csv")

# Quneitra
# Horizontal distance
qnrtdistance <- qnrtdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
qnrtdepth <- qnrtdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
qnrtarea <- qnrtdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
qnrtangle <- qnrtdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
qnrtdistance
qnrtdepth
qnrtarea
qnrtangle

# save the results 
write_csv(qnrtdepth, "../summary_stats/qnrtdepth.csv")
write_csv(qnrtarea, "../summary_stats/qnrtarea.csv")
write_csv(qnrtangle, "../summary_stats/qnrtangle.csv")
write_csv(qnrtdistance, "../summary_stats/qnrtdistance.csv")

# Qafzeh
# Horizontal distance
qzdistance <- qzdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
qzdepth <- qzdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
qzarea <- qzdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
qzangle <- qzdb %>% group_by(aoi) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
qzdistance
qzdepth
qzarea
qzangle

# save the results 
write_csv(qzdepth, "../summary_stats/qzdepth.csv")
write_csv(qzarea, "../summary_stats/qzarea.csv")
write_csv(qzangle, "../summary_stats/qzangle.csv")
write_csv(qzdistance, "../summary_stats/qzdistance.csv")

# summarise data by artefact ("site") for each measured variabled

# Horizontal distance
dbstatsdistance <- db %>% group_by(artefact) %>%
      summarise(
      count = n(),
      max = max(horizontaldistance, na.rm = TRUE),
      min = min (horizontaldistance, na.rm = TRUE),
      mean = mean(horizontaldistance, na.rm = TRUE),
      sd = sd(horizontaldistance, na.rm = TRUE),
      median = median(horizontaldistance, na.rm = TRUE),
  ) 

# Depth
dbstatsdepth <- db %>% group_by(artefact) %>%
      summarise(
      count = n(),
      max = max(maximumdepth, na.rm = TRUE),
      min = min (maximumdepth, na.rm = TRUE),
      mean = mean(maximumdepth, na.rm = TRUE),
      sd = sd(maximumdepth, na.rm = TRUE),
      median = median(maximumdepth, na.rm = TRUE),
  ) 

# Area
dbstatsarea <- db %>% group_by(artefact) %>%
      summarise(
      count = n(),
      max = max(holearea, na.rm = TRUE),
      min = min (holearea, na.rm = TRUE),
      mean = mean(holearea, na.rm = TRUE),
      sd = sd(holearea, na.rm = TRUE),
      median = median(holearea, na.rm = TRUE),
  ) 

# Angle
dbstatsangle <- db %>% group_by(artefact) %>%
      summarise(
      count = n(),
      max = max(angle, na.rm = TRUE),
      min = min (angle, na.rm = TRUE),
      mean = mean(angle, na.rm = TRUE),
      sd = sd(angle, na.rm = TRUE),
      median = median(angle, na.rm = TRUE),
  ) 

# see data summary
dbstatsdistance
dbstatsdepth
dbstatsarea
dbstatsangle

# save the results 
write_csv(dbstatsdepth, "../summary_stats/datastatsdepth.csv")
write_csv(dbstatsarea, "../summary_stats/datastatsarea.csv")
write_csv(dbstatsangle, "../summary_stats/datastatsangle.csv")
write_csv(dbstatsdistance, "../summary_stats/datastatsdistance.csv")

# General inventory (counts) table organised by site and aoi
inventory <- crosstable(db, c(artefact), by=c(aoi), total = "both")

print(inventory)
write_csv(inventory, "../summary_stats/inventory.csv")

```

# Plot data
```{r}

# converting to categorical
db$aoi = as.factor(db$aoi) 
db$incision = as.factor(db$incision) 
db$profile = as.factor(db$profile) 

# Boxplot for visualise distance, depth, area and angle data organised by AOI and Site

# Angle
angle <- ggplot(data = db, aes(x = artefact, y = angle, fill = aoi, col = artefact)) +  
            geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Angle (º)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.2, position=position_jitter(w=0.1,h=0.1))

print(angle)

ggsave("../plots/angle.png", dpi = 300)


# Depth
depth <- ggplot(data = db, aes(x = artefact, y = maximumdepth, fill = aoi, col = artefact)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Maximum depth (μm)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.2, position=position_jitter(w=0.1,h=0.1)) 

print(depth)

ggsave("../plots/depth.png", dpi = 300)

# Distance
distance <- ggplot(data = db, aes(x = artefact, y = horizontaldistance, fill = aoi, col = artefact)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Horizontal distance (μm)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.2, position=position_jitter(w=0.1,h=0.1))

print(distance)

ggsave("../plots/distance.png")

# Area
area <- ggplot(data = db, aes(x = artefact, y = holearea, fill = aoi, col = artefact)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Area of the cut (μm2)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.2, position=position_jitter(w=0.1,h=0.1)) 

print(area)

ggsave("../plots/area.png")

a <- ggarrange(angle, depth, distance, area, ncol = 2, nrow = 2, common.legend = T)

print(a)
ggsave("../plots/combinedAOI.png", dpi = 300)

# Boxplot for visualise distance, depth, area and angle data, together and organised by Site

# Angle
anglesite <- ggplot(data = db, aes(x = artefact, y = angle)) +  
            geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Angle (º)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.8, position=position_jitter(w=0.1,h=0.1))

print(anglesite)

ggsave("../plots/anglesite.png", dpi = 300)


# Depth
depthsite <- ggplot(data = db, aes(x = artefact, y = maximumdepth,)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Maximum depth (μm)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.8, position=position_jitter(w=0.1,h=0.1)) 

print(depthsite)

ggsave("../plots/depthsite.png", dpi = 300)

# Distance
distancesite <- ggplot(data = db, aes(x = artefact, y = horizontaldistance)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Horizontal distance (μm)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.8, position=position_jitter(w=0.1,h=0.1))

print(distancesite)

ggsave("../plots/distancesite.png", dpi =300)

# Area
areasite <- ggplot(data = db, aes(x = artefact, y = holearea)) +  
           geom_boxplot() +
           theme_bw()+scale_fill_grey(start = 0.8, end = 1) +
           labs(x = "", y = "Area of the cut (μm2)") +
           scale_x_discrete(labels = c("Amud 1", "Amud 2", "Manot", "Quneitra", "Qafzeh")) +
           geom_jitter(aes(colour = artefact), alpha=0.2, position=position_jitter(w=0.1,h=0.1)) 

print(areasite)

ggsave("../plots/areasite.png", dpi = 300)

a <- ggarrange(anglesite, depthsite, distancesite, areasite, ncol = 2, nrow = 2, common.legend = T)

print(a)
ggsave("../plots/combined.png", dpi = 300)


```


# Anova analysis
```{r}
# one-way anova test the effect of raw material on damage

# The null hypothesis (H0) of the ANOVA is no difference (between artefacts) in means (for each geometric feature of the incisions), and the alternative hypothesis (Ha) is that the means are different from one another (between sites).

# check if there are differences among group means
## horizontal distance (width)
onewaymodelwidth <- aov(horizontaldistance ~ artefact, data = db)
summary(onewaymodelwidth)

## Depth
onewaymodeldepth <- aov(maximumdepth ~ artefact, data = db)
summary(onewaymodeldepth)

## Area of the hole
onewaymodelarea <- aov(holearea ~ artefact, data = db)
summary(onewaymodelarea)

## Angle
onewaymodelangle <- aov(angle ~ artefact, data = db)
summary(onewaymodelangle)

# The p-value of the raw material variable is low (p < 0.001), so it appears that the site (at least one) has a real impact on the features.

# check normality
hist(onewaymodelwidth$residuals)
hist(onewaymodeldepth$residuals)
hist(onewaymodelarea$residuals)
hist(onewaymodelangle$residuals)

# library(car)
qqPlot(onewaymodelwidth$residuals)
qqPlot(onewaymodeldepth$residuals)
qqPlot(onewaymodelarea$residuals)
qqPlot(onewaymodelangle$residuals)

# check if model fits the assumption, the homogeneity of variances
par(mfrow=c(2,2))
plot(onewaymodelwidth)
par(mfrow=c(1,1))

par(mfrow=c(2,2))
plot(onewaymodeldepth)
par(mfrow=c(1,1))

par(mfrow=c(2,2))
plot(onewaymodelarea)
par(mfrow=c(1,1))

par(mfrow=c(2,2))
plot(onewaymodelangle)
par(mfrow=c(1,1))

# The model fits the assumption of homoscedasticity

# ANOVA tells us if there are differences among group means (sites and features), but not where those differences are. To find out which groups are statistically different from one another, you perform a Tukey’s Honestly Significant Difference Tukey’s HSD post-hoc test for pairwise comparisons.

# check which groups are statistically different from one another
TukeyHSD(onewaymodelwidth, conf.level=.95) 
plot(TukeyHSD(onewaymodelwidth, conf.level=.95), las = 1, cex.axis=0.5)

TukeyHSD(onewaymodeldepth, conf.level=.95) 
plot(TukeyHSD(onewaymodeldepth, conf.level=.95), las = 1, cex.axis=0.5)

TukeyHSD(onewaymodelarea, conf.level=.95) 
plot(TukeyHSD(onewaymodelarea, conf.level=.95), las = 1, cex.axis=0.5)

TukeyHSD(onewaymodelangle, conf.level=.95) 
plot(TukeyHSD(onewaymodelangle, conf.level=.95), las = 1, cex.axis=0.5)

# Some observations from the post-hoc test results:
# All raw materials are significantly different from each other. All p-value (here represented as p adj) are smaller than 0.05.

```

---

End of Script

